Prehensile Manipulation Planning: Modelling, Algorithms and Implementation

### 2020-10-19 hpp.tex
Small improvements.

Robots in industrial manufacturing are today mostly programmed by
hand. They repeat the same motion thousands of times with a high
accuracy. Automating a task with some variability is however very
challenging since it requires more programming effort to integrate
sensors and some kind of motion planning in the process. The work
described in this paper is a small step to the simplification of
industrial process automation in the presence of some variability,
like the initial position of some object or the presence of unknown
obstacles. Although the work only covers the motion planning -- and
more accurately the manipulation planning -- issue, we think that it
is important not only to develop algorithms, but also to provide them
within an open-source software platform in order to make the
evaluation and then the integration of those algorithms easier.

This \paper therefore describes a software platform called \href{https://humanoid-path-planner.github.io/hpp-doc}{Humanoid Path
  Planner} taylored for manipulation planning in robotics. It can handle many types of robots, from manipulator arms to legged humanoid robots.
The main contributions are:
\begin{itemize}
\item an original and general modelling of prehensile manipulation based on
  non-linear constraints,
\item an original solver for non-linear constraints that is able to handle implicit and explicit constraints,
\item a manipulation planning algorithm that is able to tackle a great variety of manipulation planning problems,
\item an open-source software suite that implements all the above following state of the art development tools and methods.
\end{itemize}

@@

Manipulation planning is a particular instance of path planning were
some objects are moved by some robots.
Although several instances of
manipulation problem exist like manipulation by
pushing~\cite{BenRiv1998}, or by throwing~\cite{WooZacLyn2017}, as well as
multicontact planning \cite{Bretl2006}, \cite{LenVaiYosKhe2013},
\cite{TonDelPetParManMan2018}, in this paper, we are only concerned with
prehensile manipulation planning.
The configuration space of the whole system is constrained by
non-linear constraints due to the fact that objects cannot move by
themselves and should stay in a stable pose when not grasped by a
robot.  The accessible configuration space is thus a union of
submanifolds as defined in the previous section. Each of these
manifolds is moreover a foliation where each leaf corresponds to a
stable position of an object or to a grasp of an object by a
gripper. The geometrical structure of the problem is now well
understood~\cite{AlaSimLau89}.

@@

Navigation Among Movable Obstacles (NAMO) \cite{wilfong1988motion}, \cite{stilman2008planning} consists in finding
a path for a robot that needs to displace objects in order to move to a goal
configuration. The final position of the objects does not matter in this case.

Rearrangement planning \cite{ota2004rearrangement}, \cite{LIS210},
\cite{KroBek2015}, \cite{LerPha2015} consists in finding a sequence of
manipulation paths that move some objects from an initial pose to a
final pose. The final configuration of the robot is not specified. A
simplifying asumption is the existence of a monotone solution, that is
a solution where each object is grasped once only and is moved from
its initial pose to its final pose \cite{stilman2008planning},
\cite{stilman2007manipulation}, \cite{srivastavaetal2014icra}, \cite{nieuwenhuisen2008effective},
\cite{ota2004rearrangement}. They mostly use two-level methods
composed of a symbolic or task planner and of a motion
planner~\cite{cambon:hal-01976081}, \cite{KaeLoz2013}, \cite{18-toussaint-RSS}.

@@

One of the main differences between our solution and the previously cited ones resides in the way manipulation constraints are compiled into a graph. To our knowledge none of the previous solutions can handle a variety of problems as large as those exposed in Section~\ref{sec:benchmarks}

@@

This representation although well suited for fixed base manipulator
arms is not well adapted for robots with a mobile base like wheeled
mobile robots, aerial robots or legged robots, since the mobility of the base
cannot be correctly represented by translation or rotation joints.
Representing a freeflying object by three virtual translations followed by
three virtual rotations called roll, pitch and yaw is indeed a poor workaround.

@@

A \textit{kinematic chain} is a tree of \textit{joints} where each
joint represents a mobility of a rigid-body link with respect to
another link or with respect to the world reference frame. To each joint is
associated a configuration space called the \textit{joint configuration space}.
The most common joints with their respective configuration spaces are
\begin{itemize}
\item linear \textit{translation} with configuration space $\reals$,
\item bounded \textit{rotation} with configuration space  $\reals$,
\item unbounded \textit{rotation} with configuration space  $SO(2)$,
\item \textit{planar} with configuration space $SE(2)$,
\item \textit{freeflyer} with configuration space $SE(3)$.
\end{itemize}
$SO(n)$ and $SE(n)$ respectively stand for \textit{special orthogonal group} and \textit{special Euclidean group}. They represents the group of rigid-body transformations and the group of rotations in $\reals^n$.

@@

The joint configuration spaces enumerated in the previous paragraph: $\mathbb{R}^n$, $SE(n)$ are all Lie groups. The group operation is $+$ for $\mathbb{R}^n$, and composition denoted as $"."$ for $SE(n)$. We refer to~\cite{LMS94} -- Appendix A for a thorough definition of Lie groups. We expose here only properties that are useful for the sequel of the \paper.

For any Lie group $\liegroup$ with neutral element $\neutral$, the tangent space at the neutral element $T_{\neutral}\liegroup$ of the group naturally maps to the tangent space at any point of the group. This means that any \textit{velocity} $\mathbf{v}\in T_{\neutral}\liegroup$ uniquely defines
\begin{enumerate}
\item a velocity $\mathbf{w}\in T_{g}\liegroup$ at any point $g$ of the group, and thus,
\item a vector field on the tangent space $T\liegroup$, and
\item by integration during unit time of the latter vector field, starting from the origin, a new point $g_{1}\in\liegroup$.
\end{enumerate}
Item 1 above is called the \textit{transport} of velocity $\mathbf{v}$ to $g$.
Item 3 above is called the exponential map of $\liegroup$ and is denoted by $\exp$.

\paragraph{Geometric interpretations}
\begin{itemize}
\item $\reals$ (and by trivial generalization $\reals^n$): the neutral element is $0$. The tangent space at 0 is isomorphic to $\reals$ and
  $$
  \forall \theta\in\reals,\  \exp(\theta) = \theta.
  $$
\item $SE(3)$: an element $g$ of $SE(3)$ can be seen as the position of a moving frame in a fixed reference frame. A point $\mathbf{x}\in\reals^3$ is mapped to $g(\mathbf{x})$. Note that $\mathbf{x}$ is also the coordinate vector of $g(\mathbf{x})$ in the moving frame $g$. If $\mathbf{v}$, $\omega$ are a linear and an angular velocities at the origin, $(\mathbf{v},\omega)$ is transported to $g$
  as the same linear and angular velocities expressed in the moving frame. In other words, if
  \begin{equation}\label{eq:homogeneous matrix}
  M=\left(\begin{array}{ll} R & \mathbf{t}\\ 0 & 1\end{array}\right)
    \end{equation}
  with $R\in SO(3)$ and $\mathbf{t}\in\reals^3$  is the homogeneous matrix
  representing $g$, and $(\mathbf{v},\mathbf{\omega})$ is a velocity in $T_{I_3}SE(3)$, the velocity transported to $g$ corresponds to the linear velocity $R\mathbf{v}$ and $R\omega$ of the moving frame.
  Integral curves of the vector field mentioned in item 2 above correspond to helicoidal motions of constant velocity expressed in the moving frame.
\end{itemize}
$SE(2)$, $SO(3)$, and $SO(2)$ are (or are isomorphic to) subgroups of $SE(3)$ and follow the same geometrical interpretation.

@@

Therefore elements of $SE(3)$ are represented by a vector in $\reals^7$ where the 3 first components represent the image of the origin (vector $\mathbf{t}$ in Equation~\ref{eq:homogeneous matrix}), the 4 last components represent a unit quaternion.

Elements of $SO(3)$ are represented by a unit vector of dimension 4.

@@

As expressed earlier, following a constant velocity $\dot{\conf}$ from the neutral element of a joint configuration space leads to another configuration denoted as
$$
\conf = \exp(\dconf).
$$
In some cases, we may specify in subscript the Lie group that is used: \href{https://github.com/stack-of-tasks/pinocchio/blob/f8f3b9a24eab527df79650e3dc73410f9a46a2b2/src/spatial/explog.hpp#L34}{$\exp_{SO(3)}$}, \href{https://github.com/stack-of-tasks/pinocchio/blob/f8f3b9a24eab527df79650e3dc73410f9a46a2b2/src/spatial/explog.hpp#L236}{$\exp_{SE(3)}$}.

@@

Following a constant velocity $\dconf\in T_{\neutral}\liegroup$ starting from $\conf_0\in\liegroup$, leads to
$$
\conf_1 = q_0.\exp(\dconf).
$$
Note that if $\liegroup=\reals$, we write
$$
\conf_1 = \conf_0 + \dconf,
$$
since the Lie group operator of $\reals$ is $+$ and $\exp_{\reals}$ is the identity.
In order to homogeneize notation, we define the following operators:
\begin{eqnarray}\label{eq:plus}
  \conf_0 \oplus \dconf &\triangleq& \conf_0.\exp(\dconf),\\
  \label{eq:minus}
  \conf_1 \ominus \conf_0 &\triangleq& \log(\conf_0^{-1}.\conf_1).
\end{eqnarray}

@@

Given a kinematic chain with joints $(J_1, \cdots, J_{njoints})$, ordered in such
a way that each joint has an index bigger than its parent in the tree, the configuration space of the robot is the cartesian product of the joint configuration spaces.
$$
\CS \triangleq \CS_{J_1}\times \cdots \times \CS_{J_{njoints}}.
$$
$\CS$ naturally inherits the Lie group structure of the joint configuration spaces through the Cartesian product.
We denote by $nq_i$, $nv_i$ the sizes of the configuration and velocity vector representations of joint $J_i$, as defined in Table~\ref{tab:vector representation}.
The configuration and velocity of the robot can thus be represented by vectors of size $nq$ and $nv$ such that
$$
nq = \sum_{i=1}^{njoints} nq_i, \ \ \ nv = \sum_{i=1}^{njoints} nv_i
$$
We denote by $iq_i$, and $iv_i$ the indices of joint $i$ in the robot configuration and velocity vectors.
$$
iq_i = \sum_{j=1}^{i-1} nq_j\ \ \ \ iv_i = \sum_{j=1}^{i-1} nv_j
$$
With these definitions and notation, the linear interpolation between two robot
configurations $\conf_0$ and $\conf_1$ is naturally written:
$$
\conf(t) = \conf_0 \oplus t (\conf_1 - \conf_0)
$$
This formula generalizes the linear interpolation to robots with freeflying bases, getting rid of singularities of roll -- pitch -- yaw parameterization.
%% begin impl
Cartesian products of Lie groups are represented by class \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupSpace.html}{\texttt{LiegroupSpace}}. Elements of these spaces are represented by classes
\begin{itemize}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElement}}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElementRef}}, and
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElementConstRef}}.
  \end{itemize}
%% end impl

@@

\begin{definition}\textbf{Non-linear constraint.} A non linear constraint is defined by a differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and is written
\begin{equation}\label{eq:constraint}
h(\conf) = 0.
\end{equation}
\end{definition}
If the robot is subject to several numerical constraints, $h_1,\cdots,h_k$ with values in $\reals^{m_1}\cdots\reals^{m_k}$, these constraints are equivalent to a single constraint $h$ with values in $\reals^{m}$, where $m=\sum_{i=1}^k m_i$, such that
$$
h(\conf) \triangleq \left(\begin{array}{c} h_1(\conf) \\ \vdots \\ h_k(\conf)
\end{array}\right).
$$
It may be useful to use a non zero right hand side for the same function $h$. We define for that parameterized non-linear constraints.
\begin{definition}\textbf{Parameterized non-linear constraints.}
  A parameterized non-linear constraint is defined by a differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and by a vector $\mathbf{h}_0$ of $\reals^m$ and is written
  $$
  h(\conf) = \mathbf{h}_0.
  $$
\end{definition}
%% begin impl
Differentiable functions are represented by class\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1DifferentiableFunction.html}{\texttt{DifferentiableFunction}}.
%% end impl

@@

In this \paper, we will make use of the term Jacobian in a generalized way.
If $h$ is a differentiable function from a Lie group $\liegroup_1$ to a Lie group $\liegroup_2$, and $\conf_1$ an element of $\liegroup_1$, we will denote by $\frac{\partial h}{\partial\conf}(\conf_1)$ the operator that maps velocities in $T_{\conf_1}\liegroup_1$ to the corresponding velocity of the image by $h$ in $T_{h(\conf_1)}\liegroup_2$.

@@

It is sometimes useful to produce a configuration $\conf$ that satisfies a constraint (or a set of constraints) of type~(\ref{eq:constraint}) from a configuration $\conf_0$ that does not. This action is called the \textit{projection of $\conf_0$ onto the submanifold defined by the constraint} and is performed by a numerical solver that iteratively linearizes the constraint as follows:
$$
h(\conf_{i+1}) \approx h(\conf_i) + \frac{\partial h}{\partial \conf}(\conf_{i})(\conf_{i+1}-\conf_i) = 0
$$
Iterate $\conf_{i+1}$ is computed as follows:
\begin{equation}\label{eq:HierarchicalIterative}
\conf_{i+1} = \conf_i-\alpha_i\frac{\partial h}{\partial \conf}^{+}(\conf_{i}) h(\conf_i)
\end{equation}
where $.^{+}$ denotes the Moore Penrose pseudo inverse, and $\alpha_i$ is a positive real number called the step size. Taking $\alpha_i=1$ exactly solves the linear approximation, but it may not be the best choice in general.

@@

In manipulation planning applications, where robots manipulate objects, once an object is grasped, the position of the object can be computed explicitely from the configuration of the robot. In this case, some configuration variables of the system depends on other:
$$
\conf=(\conf_{rob},\conf_{obj})\in\CS,\ \ \ \ \conf_{obj} = g_{grasp}(\conf{rob}).
$$
Although this constraint may fit definition~(\ref{eq:constraint}) by defining
\begin{equation}\label{eq:explicit-grasp}
h(\conf) \triangleq \conf_{obj} \ominus g_{grasp}(\conf_{rob}),
\end{equation}
solving this constraint possibly with other constraints using an iterative scheme~(\ref{eq:HierarchicalIterative}) is obviously sub-optimal.
Notice that the right hand side of~(\ref{eq:explicit-grasp}) is the $\log$ of the pose of the object when grasped, with respect to the actual pose of the object.

@@

% begin impl
The solver by substitution described in this section is implemented by class \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1solver_1_1BySubstitution.html}{\texttt{SolverBySubstitution}}, that store an instance of \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1ExplicitConstraintSet.html}{\texttt{ExplicitConstraintSet}}.

\paragraph{Important remark} As mentioned in Table~\ref{tab:vector representation}, the configuration and velocity vectors may have different sizes. As a consequence, index sets $in$ and $out$ in Definition~\ref{def:Explicit} correspond to configuration vector indices, while in Expression~(\ref{eq:jacobian-explicit}) they correspond to velocity vector indices. To keep notation simple, we use the same notation for different sets.

@@

Then the corresponding constrained path $\tilde{P}$ is defined on the same interval by
$$
\forall t\in[0,T],\ \ \ \tilde{P}(t) = \mathbf{proj}(P(t))
$$
Paths are implemented by class \href{https://gepettoweb.laas.fr/hpp/hpp-core/doxygen-html/classhpp_1_1core_1_1Path.html}{\texttt{Path}}. Several implementations of unconstrained paths are provided:\\
\href{https://gepettoweb.laas.fr/hpp/hpp-core/doxygen-html/classhpp_1_1core_1_1StraightPath.html}{\texttt{StraightPath}} for linear interpolation generalized to Lie groups, \href{https://gepettoweb.laas.fr/hpp/hpp-core/doxygen-html/classhpp_1_1core_1_1ReedsSheppPath.html}{\texttt{ReedsSheppPath}}, \href{https://gepettoweb.laas.fr/hpp/hpp-core/doxygen-html/classhpp_1_1core_1_1DubinsPath.html}{\texttt{DubinsPath}} for non\-holo\-nomic mobile robots.

@@

Projecting configurations at path evaluation has the advantage of not losing information. In return, the projection of a continuous path may be discontinuous. Before inserting a projected path in a roadmap for instance, it is therefore necessary to detect possible discontinuities. \cite{mirabel:hal-01360409} proposes two algorithms to check whether a projected path is continuous.

@@

Note that when a grasp constraint and its complement are combined, they
constitute an explicit constraint since the pose of the object grasped
uniquely depends on the configuration of the robot that grasps.

>>>>

Robots in industrial manufacturing are today mostly programmed by
hand. They repeat the same motion thousands of times with a high
accuracy. Automating a task with some variability is however very
challenging since it requires more programming effort to integrate
sensors and motion planning in the process. The work described in this
paper is a small step to the simplification of industrial process
automation in the presence of some variability, like the variation of
the initial position of some object or the presence of unknown
obstacles. The work only covers the motion planning -- and
more accurately the manipulation planning -- issue. The integration in a whole
process is still under development. We think that it
is important not only to develop algorithms, but also to provide them
within an open-source software platform in order to make the
evaluation and then the integration of those algorithms easier.

This \paper therefore describes a software platform called \href{https://humanoid-path-planner.github.io/hpp-doc}{Humanoid Path
  Planner} taylored for manipulation planning in robotics. It can handle many types of robots, from manipulator arms to legged humanoid robots.
The main contributions are:
\begin{itemize}
\item an original and general modelling of prehensile manipulation based on
  non-linear constraints,
\item an original solver for non-linear constraints that is able to handle implicit and explicit constraints,
\item a manipulation planning algorithm that is able to tackle a great variety of manipulation planning problems,
\item an open-source software suite that implements all the above following state of the art development tools and methods.
\item a docker image of the above software with installation instructions provided with this paper.
\end{itemize}

@@

Manipulation planning is a particular instance of path planning were
some objects are moved by some robots.
Although several instances of
manipulation problem exist like manipulation by
pushing~\cite{BenRiv1998}, or by throwing~\cite{WooZacLyn2017}, as well as
multicontact planning \cite{Bretl2006}, \cite{LenVaiYosKhe2013},
\cite{TonDelPetParManMan2018}, in this paper, we are only concerned with
prehensile manipulation.
The configuration space of the whole system is constrained by
non-linear constraints due to the fact that objects cannot move by
themselves and should stay in a stable pose when not grasped by a
robot.  The accessible configuration space is thus a union of
submanifolds as defined in the previous section. Each of these
manifolds may moreover be a foliation where each leaf corresponds to a
stable position of an object or to a grasp of an object by a
gripper. The geometrical structure of the problem has been well
understood for a long time~\cite{AlaSimLau89}.

@@

Navigation Among Movable Obstacles (NAMO) \cite{wilfong1988motion}, \cite{stilman2008planning} consists in finding
a path for a robot that needs to move objects in order to reach a goal
configuration. The final position of the objects does not matter in this case.

Rearrangement planning \cite{ota2004rearrangement}, \cite{LIS210},
\cite{KroBek2015}, \cite{LerPha2015} consists in finding a sequence of
manipulation paths that move some objects from an initial pose to a
final pose. The final configuration of the robot is not specified. A
simplifying asumption is the existence of a monotone solution, that is
a solution where each object is grasped at most once and is moved from
its initial pose to its final pose \cite{stilman2008planning},
\cite{stilman2007manipulation}, \cite{srivastavaetal2014icra}, \cite{nieuwenhuisen2008effective},
\cite{ota2004rearrangement}. They mostly use two-level methods
composed of a symbolic or task planner and of a motion
planner~\cite{cambon:hal-01976081}, \cite{KaeLoz2013}, \cite{18-toussaint-RSS}.

@@

One of the main differences between our solution and the previously cited ones resides in the way manipulation constraints are compiled into a graph. To our knowledge none of the previous solutions can handle a variety of problems as large as those exposed in Section~\ref{sec:benchmarks}.

@@

This representation although well suited for fixed base manipulator
arms is not well adapted for robots with a mobile base like wheeled
mobile robots, aerial robots or legged robots, since the mobility of the base
cannot be correctly represented by translation or rotation joints.
Representing a freeflying object by three virtual translations followed by
three virtual rotations called roll, pitch and yaw is indeed a poor workaround
due to the presence of singularities. A good illustration of this is the gimball
lock issue that arose during Apollo~13 flight. To avoid singularities, we
propose the following defintion.

@@

A \textit{kinematic chain} is a tree of \textit{joints} where each
joint represents a mobility of a rigid-body link with respect to
another link or with respect to the world reference frame. To each joint is
associated a configuration space called the \textit{joint configuration space}.
The most common joints with their respective configuration spaces are
\begin{itemize}
\item linear \textit{translation} with configuration space $\reals$,
\item bounded \textit{rotation} with configuration space  $\reals$,
\item unbounded \textit{rotation} with configuration space  $SO(2)$,
\item \textit{planar} joint with configuration space $SE(2)$,
\item \textit{freeflyer} joint with configuration space $SE(3)$.
\end{itemize}
$SO(n)$ and $SE(n)$ respectively stand for \textit{special orthogonal group} and \textit{special Euclidean group}. They represent the group of rotations and the group of rigid-body transformations in $\reals^n$.

@@

The joint configuration spaces enumerated in the previous paragraph: $\mathbb{R}^n$, $SO(n)$, and $SE(n)$ are all Lie groups. The group operation is $+$ for $\mathbb{R}^n$, and composition denoted as $"."$ for $SE(n)$. We refer to~\cite{LMS94} -- Appendix A for a thorough definition of Lie groups. We expose here only properties that are useful for the following developments.

For any Lie group $\liegroup$ with neutral element $\neutral$, the tangent space at the neutral element $T_{\neutral}\liegroup$ of the group naturally maps to the tangent space at any point of the group. This means that any \textit{velocity} $\mathbf{v}\in T_{\neutral}\liegroup$ uniquely defines
\begin{enumerate}
\item a velocity $\mathbf{w}\in T_{g}\liegroup$ at any point $g$ of the group, and thus,
\item a vector field on the tangent space $T\liegroup$, and
\item by integration during unit time of the latter vector field, starting from the origin, a new point $g_{1}\in\liegroup$.
\end{enumerate}
Item 1 above is called the \textit{transport} of velocity $\mathbf{v}$ to $g$.
Item 3 is called the exponential map of $\liegroup$ and is denoted by $\exp$.

\paragraph{Geometric interpretations}
\begin{itemize}
\item $\reals$ (and by trivial generalization $\reals^n$): the neutral element is $0$. The tangent space at 0 is isomorphic to $\reals$ and
  $$
  \forall \theta\in\reals,\  \exp(\theta) = \theta.
  $$
\item $SE(3)$: an element $g$ of $SE(3)$ can be seen as the position of a moving frame in a fixed reference frame. A point $\mathbf{x}\in\reals^3$ is mapped to $g(\mathbf{x})$. Note that $\mathbf{x}$ is also the coordinate vector of $g(\mathbf{x})$ in the moving frame $g$. If $\mathbf{v}$, $\omega$ are a linear and an angular velocities at the origin, $(\mathbf{v},\omega)$ is \textit{transported} to $g$
  as the same linear and angular velocities expressed in the moving frame. In other words, if
  \begin{equation}\label{eq:homogeneous matrix}
  M=\left(\begin{array}{ll} R & \mathbf{t}\\ 0 & 1\end{array}\right)
    \end{equation}
  with $R\in SO(3)$ and $\mathbf{t}\in\reals^3$  is the homogeneous matrix
  representing $g$, and $(\mathbf{v},\mathbf{\omega})$ is a velocity in $T_{I_3}SE(3)$, the velocity \textit{transported} to $g$ corresponds to linear and angular velocities $R\mathbf{v}$ and $R\omega$ of the moving frame.
  Integral curves of the vector field mentioned in item 2 above correspond to helicoidal motions of constant velocity expressed in the moving frame.
\end{itemize}
$SE(2)$, $SO(3)$, and $SO(2)$ are (or are isomorphic to) subgroups of $SE(3)$ and follow the same geometrical interpretation.

@@

Therefore elements of $SE(3)$ are represented by a vector in $\reals^7$ where the 3 first components represent the image of the origin (vector $\mathbf{t}$ in Equation~\ref{eq:homogeneous matrix}), the 4 last components $(x,y,z,w)$ represent unit quaternion $w + x {i} + y {j} + z {k}$.

Elements of $SO(3)$ are likewise represented by a unit vector of dimension 4.

@@

As expressed earlier, following a constant velocity\footnote{More precisely, following the vector field generated by $\dot{\conf}\in T_{\neutral}\liegroup$ according to the Lie group structure} $\dot{\conf}$ from the neutral element of a joint configuration space leads to another configuration denoted as
$$
\conf = \exp(\dconf).
$$
In some cases, we may specify in subscript the Lie group that is used: \href{https://github.com/stack-of-tasks/pinocchio/blob/f8f3b9a24eab527df79650e3dc73410f9a46a2b2/src/spatial/explog.hpp#L34}{$\exp_{SO(3)}$}, \href{https://github.com/stack-of-tasks/pinocchio/blob/f8f3b9a24eab527df79650e3dc73410f9a46a2b2/src/spatial/explog.hpp#L236}{$\exp_{SE(3)}$}.

@@

Following a constant velocity $\dconf\in T_{\neutral}\liegroup$ starting from $\conf_0\in\liegroup$, leads to
$$
\conf_1 = q_0.\exp(\dconf).
$$
Note that if $\liegroup=\reals$, we write
$$
\conf_1 = \conf_0 + \dconf,
$$
since the Lie group operator of $\reals$ is $+$ and $\exp_{\reals}$ is the identity.
In order to homogeneize notation, we define the following operators. For any
$\conf_0,\conf_1\in\liegroup$ and $\dconf\in T_{\neutral}\liegroup$:
\begin{eqnarray}\label{eq:plus}
  \conf_0 \oplus \dconf &\triangleq& \conf_0.\exp(\dconf) \in \liegroup,\\
  \label{eq:minus}
  \conf_1 \ominus \conf_0 &\triangleq& \log(\conf_0^{-1}.\conf_1) \in T_{\neutral}\liegroup.
\end{eqnarray}

@@

Given a kinematic chain with joints $(J_1, \cdots, J_{njoints})$, ordered in such
a way that each joint has an index bigger than its parent in the tree, the configuration space of the robot is the cartesian product of the joint configuration spaces.
$$
\CS \triangleq \CS_{J_1}\times \cdots \times \CS_{J_{njoints}}.
$$
$\CS$ naturally inherits the Lie group structure of the joint configuration spaces through the Cartesian product.
We denote by $nq_i$, $nv_i$ the sizes of the configuration and velocity vector representations of joint $J_i$, as defined in Table~\ref{tab:vector representation}.
The configuration and velocity of the robot can thus be represented by vectors of size $nq$ and $nv$ such that
$$
nq = \sum_{i=1}^{njoints} nq_i, \ \ \ nv = \sum_{i=1}^{njoints} nv_i
$$
We denote by $iq_i$, and $iv_i$ the indices of joint $i$ in the robot configuration and velocity vectors.
$$
iq_i = \sum_{j=1}^{i-1} nq_j\ \ \ \ iv_i = \sum_{j=1}^{i-1} nv_j
$$
With these definitions and notation, the linear interpolation between two robot
configurations $\conf_0$ and $\conf_1$ is naturally written:
$$
\conf(t) = \conf_0 \oplus t (\conf_1 - \conf_0)
$$
This formula generalizes the linear interpolation to robots with freeflying bases, getting rid of singularities of roll -- pitch -- yaw parameterization.
%% begin impl
Cartesian products of Lie groups are represented by Class \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupSpace.html}{\texttt{LiegroupSpace}}. Elements of these spaces are represented by classes
\begin{itemize}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElement}}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElementRef}}, and
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElementConstRef}}.
  \end{itemize}
%% end impl

@@

\begin{definition}\textbf{Non-linear constraint.} A non linear constraint is defined by a differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and is written
\begin{equation}\label{eq:constraint}
h(\conf) = 0.
\end{equation}
\end{definition}
If the robot is subject to several numerical constraints, $h_1,\cdots,h_k$ with values in $\reals^{m_1}\cdots\reals^{m_k}$, these constraints are equivalent to a single constraint $h$ with values in $\reals^{m}$, where $m=\sum_{i=1}^k m_i$, such that
$$
h(\conf) \triangleq \left(\begin{array}{c} h_1(\conf) \\ \vdots \\ h_k(\conf)
\end{array}\right).
$$
It may be useful to use a non zero right hand side for the same function $h$. We define for that parameterized non-linear constraints.
\begin{definition}\textbf{Parameterized non-linear constraint.}
  A parameterized non-linear constraint is defined by a differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and by a vector $\mathbf{h}_0$ of $\reals^m$ and is written
  $$
  h(\conf) = \mathbf{h}_0.
  $$
\end{definition}
%% begin impl
Differentiable mappings are represented by abstract Class\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1DifferentiableFunction.html}{\texttt{DifferentiableFunction}}.
%% end impl

@@

In this \paper, we will make use of the term Jacobian in a generalized way.
If $h$ is a differentiable function from a Lie group $\liegroup_1$ to a Lie group $\liegroup_2$, and $\conf_1$ an element of $\liegroup_1$, we will denote by $\frac{\partial h}{\partial\conf}(\conf_1)$ the operator that maps velocities in $T_{\conf_1}\liegroup_1$ to the velocity in $T_{h(\conf_1)}\liegroup_2$ \textit{transported by} $h$\footnote{If $\dconf\in T_{\conf_1}\liegroup_1$ is a velocity along a time parameteriezd curve $\gamma$, $\frac{\partial h}{\partial\conf}(\conf_1)\dconf$ is the velocity at
  the same time along Curve $h(\gamma)$.}.

@@

It is sometimes useful to produce a configuration $\conf$ that satisfies a constraint (or a set of constraints) of type~(\ref{eq:constraint}) from a configuration $\conf_0$ that does not. This action is called the \textit{projection of $\conf_0$ onto the submanifold defined by the constraint} and is performed by a numerical solver that iteratively linearizes the constraint as follows:
$$
h(\conf_{i+1}) \approx h(\conf_i) + \frac{\partial h}{\partial \conf}(\conf_{i})(\conf_{i+1}-\conf_i) = 0
$$
Iterate $\conf_{i+1}$ is computed as follows:
\begin{equation}\label{eq:HierarchicalIterative}
\conf_{i+1} = \conf_i-\alpha_i\frac{\partial h}{\partial \conf}^{+}(\conf_{i}) h(\conf_i)
\end{equation}
where $.^{+}$ denotes the Moore Penrose\footnote{He has just been awarded the Nobel Prize.} pseudo inverse, and $\alpha_i$ is a positive real number called the step size. Taking $\alpha_i=1$ exactly solves the linear approximation, but it may not be the best choice in general.

@@

In manipulation planning applications, where robots manipulate objects, once an object is grasped, the position of the object can be computed explicitely from the configuration of the robot. In this case, some configuration variables of the system depend on other:
$$
\conf=(\conf_{rob},\conf_{obj})\in\CS,\ \ \ \ \conf_{obj} = g_{grasp}(\conf{rob}).
$$
Although this constraint may fit definition~(\ref{eq:constraint}) by defining
\begin{equation}\label{eq:explicit-grasp}
h(\conf) \triangleq \conf_{obj} \ominus g_{grasp}(\conf_{rob}),
\end{equation}
solving this constraint possibly with other constraints using an iterative scheme~(\ref{eq:HierarchicalIterative}) is obviously sub-optimal.
Notice that the right hand side of~(\ref{eq:explicit-grasp}) is the $\log$ of the pose of the object when grasped, with respect to the actual pose of the object.

@@

% begin impl
The solver by substitution described in this section is implemented by Class \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1solver_1_1BySubstitution.html}{\texttt{SolverBySubstitution}}, that store an instance of \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1ExplicitConstraintSet.html}{\texttt{ExplicitConstraintSet}}.

\paragraph{Important remark} As mentioned in Table~\ref{tab:vector representation}, the configuration and velocity vectors may have different sizes. As a consequence, index sets $in$ and $out$ in Definition~\ref{def:Explicit} correspond to configuration vector indices, while in Expression~(\ref{eq:jacobian-explicit}), they correspond to velocity vector indices. To keep notation simple, we use the same notation for different sets.

@@

Then the corresponding constrained path $\tilde{P}$ is defined on the same interval by
$$
\forall t\in[0,T],\ \ \ \tilde{P}(t) = \mathbf{proj}(P(t))
$$
Paths are implemented by Class \href{https://gepettoweb.laas.fr/hpp/hpp-core/doxygen-html/classhpp_1_1core_1_1Path.html}{\texttt{Path}}. Several implementations of unconstrained paths are provided:\\
\href{https://gepettoweb.laas.fr/hpp/hpp-core/doxygen-html/classhpp_1_1core_1_1StraightPath.html}{\texttt{StraightPath}} for linear interpolation generalized to Lie groups, \href{https://gepettoweb.laas.fr/hpp/hpp-core/doxygen-html/classhpp_1_1core_1_1ReedsSheppPath.html}{\texttt{ReedsSheppPath}}, \href{https://gepettoweb.laas.fr/hpp/hpp-core/doxygen-html/classhpp_1_1core_1_1DubinsPath.html}{\texttt{DubinsPath}} for non\-holo\-nomic mobile robots.

@@

Projecting configurations at path evaluation has the advantage of not losing information. In return, the projection of a continuous path may be discontinuous. Before inserting a projected path in a roadmap for instance, it is therefore necessary to detect possible discontinuities. \cite{Hauser-RSS-13} proposes a solution to this problem. \cite{mirabel:hal-01360409} describe two algorithms to check whether a projected path is continuous. These algorithms are implemented by
classes \href{https://gepettoweb.laas.fr/hpp/hpp-core/doxygen-html/classhpp_1_1core_1_1pathProjector_1_1Dichotomy.html}{\texttt{pathProjector::Dichotomy}} and \href{https://gepettoweb.laas.fr/hpp/hpp-core/doxygen-html/classhpp_1_1core_1_1pathProjector_1_1Progressive.html}{\texttt{pathProjector::Progressive}}.

@@

Note that when a grasp constraint and its complement are combined, they
constitute an explicit constraint since the pose of the object grasped
uniquely depends on the configuration of the robot that grasps the object.
